---
title: "Impact of Redlining In LA On Avian Biodiversity"
subtitle: "Exploring patterns of environmental justice"
author: Amanda G. Overbye
date: last-modified
execute: 
  eval: true
  warning: false
format:
  html:
    code-fold: true
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
# Load in libraries
library(here)
library(tidyverse)
library(sf)
library(tmap)
library(ggplot2)
library(cowplot)
library(knitr)
```

## Part 1: Legacy of redlining in current environmental (in)justice

```{r}
# Create color palette for this project
robinhood1_palatte <-c("#698847ff","#519e99ff", "#e3ba11ff", "#8c241cff", "#8f592aff")

```

```{r, message=FALSE, results='hide', warning=FALSE}
# Read in data

# Read in ejscreen
ejscreen <- sf::st_read(here::here("data","ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))

# Filter California from ejscreen
cali <- ejscreen %>%
  dplyr::filter(ST_ABBREV == "CA")

# Filter Los Angeles County from cali
la_county <- cali %>%
  dplyr::filter(CNTY_NAME %in% c("Los Angeles County"))

# Read in mapping inequality 
la_map_ineq <- st_read(here::here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))

# Read in bird data 
bird_obs <- st_read(here::here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))

```

```{r}
# Create a bbox for la_map_ineq
ineq_bbox <- st_bbox(la_map_ineq)
```

```{r}
# Create map of redlining in LA

tm_shape(cali, bbox = ineq_bbox) +
  tm_polygons(col = "#c8bfa619", alpha = .4) +
tm_shape(la_county) +
  tm_polygons(col = "#c8bfa619", alpha = .4) +
tm_shape(la_map_ineq) +
  tm_polygons(col = "grade", 
             palette = c("#698847ff", "#519e99ff","#e3ba11ff", "#8c241cff", "#8f592aff"),
                         title = "HOLC Grade") +
   tm_scale_bar(position = c("right", "top")) +
   tm_compass( show.labels = 2, 
              type = "arrow",
              size = 1, 
              text.size = .75,
              position = c("right", "top")) + 
   tm_layout(main.title = "HOLC Grade Areas in LA County")
```

```{r}
# Create function to check if la_county and la_map_ineq have same crs system
check_crs <- function(la_county, la_map_ineq) {
  if (!st_crs(la_county) == st_crs(la_map_ineq)) {
    warning("CRS do not match. Use st_transform to creating match crs systems.")
  } else {
    print("CRS match!")
  }
}
```

```{r, message=FALSE, results='hide', warning=FALSE}
# Check to see if la_county and la_map_ineq have same crs
check_crs(la_county, la_map_ineq)
```

```{r}
# Ensure la_map_ineq and la_county are using the same la_map_ineq crs system
la_map_ineq <- st_transform(la_map_ineq, crs = st_crs(la_county))
```

```{r}
# Check to see if la_county and la_map_ineq have same crs again
check_crs(la_county, la_map_ineq)
```

```{r}
# Join la_county and la_map_ineq with st_intersects()
joined_la <- st_join(la_county, la_map_ineq, join = st_intersects)
```

```{r}
# Create table of percent in each HOLC group
HOLC_percent <- joined_la %>%
  group_by(grade) %>%
  summarise(count = n()) %>%
  mutate(HOLC_percent = (count / sum(count)) * 100)
  
```

```{r}
# Get mean values for percentage low income, PM2.5 percentiles, and low life expediency percentiles
sum_current_conditions <- joined_la %>%
  group_by(grade) %>%
  summarise( 
    low_income_mean = (mean(LOWINCPCT, na.rm = TRUE) * 100),
    pm2_mean = mean(P_PM25, na.rm = TRUE),
    life_exp_mean = mean(P_LIFEEXPPCT, na.rm = TRUE))
```

```{r}
# Drop geometry column for table
sum_current_conditions_table <- st_drop_geometry(sum_current_conditions)
```

```{r}
kable(sum_current_conditions_table, col.names = c("HOLC Grade", "Low Income Percentage", "PM2.5 Mean Percentile", "Low Life Expectancy Percentile"))
```

```{r}
# Create graph for sum_current_conditions
ggplot(data = sum_current_conditions, aes(x = grade, y = pm2_mean, fill = grade)) +
  geom_col() +
   scale_fill_manual(values=robinhood1_palatte) +
  labs(title = "Mean PM 2.5 Percentile by HOLC Grade", 
       x = "HOLC Grade", 
       y = "Mean Life PM 2.5 Percentile")

ggplot(data = sum_current_conditions, aes(x = grade, y = low_income_mean, fill = grade)) +
  geom_col() +
   scale_fill_manual(values=robinhood1_palatte) +
  labs(title = "Mean Low Income Percentile by HOLC Grade", 
       x = "HOLC Grade", 
       y = "Mean Percent Low Income")

ggplot(data = sum_current_conditions, aes(x = grade, y = life_exp_mean, fill = grade)) +
  geom_col() +
   scale_fill_manual(values=robinhood1_palatte) +
  labs(title = "Mean Life Expectancy Percentile by HOLC Grade", 
       x = "HOLC Grade", 
       y = "Mean Life Expectancy Percentile")
```

### Looking At The Data

When looking at the geographical map of Los Angeles County, we can see that many of the grade C and D HOLC areas are clustered in the center of the city, whereas the A and B grade areas are more prominent on the outskirts. The tables and graphs on income level, particulate matter, and life expectancy show that areas affected by redlining have lower incomes, more particulate matter in the air, and lower life expectancies. These findings illustrate that redlining has had a negative impact on those living in these areas

## Part 2: Legacy of redlining in biodiversity observations

```{r}
# Create function to check if la_county and bird_obs have same crs system
check_crs2 <- function(la_county, bird_obs) {
  if (!st_crs(la_county) == st_crs(bird_obs)) {
    warning("CRS do not match. Use st_transform to creating match crs systems.")
  } else {
    print("CRS match")
  }
}
```

```{r, message=FALSE, results='hide', warning=FALSE}
# Check to see if la_county and bird_obs have same crs
check_crs2(la_county, bird_obs)
```

```{r}
# st_transform so bird_obs and la_county have same crs
bird_obs <- st_transform(bird_obs, crs = st_crs(la_county))
```

```{r}
# Check to see if la_county and bird_obs have same crs again
check_crs2(la_county, bird_obs)
```

```{r}
la_birds <- st_join(la_map_ineq, bird_obs, join = st_intersects)
```

```{r}
# Create table of bird observations by HOLC grade
birds_obs_percent <- la_birds %>%
  group_by(grade) %>%
  summarise(obs_count = n()) %>%
  mutate(percentage = (obs_count / sum(obs_count)) * 100)
```

```{r}
# Create col chart of the percent of birds in each grade
ggplot(data = birds_obs_percent, aes(x = grade, y = percentage, fill = grade)) +
  geom_col() +
  scale_fill_manual(values=robinhood1_palatte) +
  labs(title = "Percentage of Bird Observations by HOLC Grade", 
       x = "HOLC Grade", 
       y = "Percentage of Bird Observations")
```

### Looking At The Data

Our data show that the majority of bird observations have been done within the C HOLC grade area. The second most common grades are the A and D grade areas. This is unexpected because previous data has shown that there are fewer bird observations in heavily redlined areas (Ellis-Soto et al., 2023). This indicates that there may be a more complex relationship between bird observations and redlining that could be investigated in the future.

**Citations:**

Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical redlining is associated with increasing geographical disparities in bird biodiversity sampling in the United States. Nature Human Behaviour, 1-9

GBIF.org, gbif-birds-LA dataset

Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan Connolly, et al., “Mapping Inequality,” American Panorama, ed.Robert K. Nelson and Edward L. Ayers, accessed October 24, 2023

U.S. Environmental Protection Agency (EPA), 2024. EJScreen Technical Documentation, accessed October 24, 2023
